Modelling the emergence of whisker barrels

### 2019-12-19 paper/briefpaper/brief.tex
sfs

\begin{abstract}
Brain development relies on an interplay of genetic specification and self-organisation. Striking examples of this interplay can be found in the rodent somatosensory brainstem, thalamus, and cortex, where the arrangement of the facial whiskers is preserved in the arrangement of cell aggregates to form precise somatotopic maps. We show in simulation how realistic whisker maps can self-organize, by assuming that information is exchanged between adjacent cells only, under the guidance of gene expression gradients. The resulting model provides a simple account of how genetic information can constrain spontaneous pattern formation to faithfully reproduce functional maps in subsequent brain structures.
\end{abstract}

@@

\dropcap{S}patial patterns in neural connectivity provide clues about the constraints under which brains evolve and develop \citep{purves_iterated_1992}. Perhaps the most distinctive pattern can be found in the rodent barrel cortex \cite{woolsey_structural_1970}. The barrels are large cortical columns, identifiable from birth in layer 4 of primary somatosensory cortex as dense clusters of thalamocortical axons, which are enclosed by borders a few neurons thick from postnatal day 3 \citep{erzurumlu_development_2012}. In the plane tangential to the cortical surface the barrels constitute a somatotopic map of the whiskers, with cells of adjacent barrels responding most strongly and quickly to deflection of adjacent whiskers \citep{armstrong-james_flow_1992}. Barrel patterning reflects subcortical whisker maps comprising cell aggregates called barrelettes in the brainstem and barreloids in the thalamus.
%\cite{van_der_loos_barreloids_1976}.

@@

The present results suggest that the key requirements for the emergence of realistic barrel patterning are i) at each cortical location thalamocortical projections compete for a limited number of available synaptic connections (Eqs.~\ref{eq:dc}--\ref{eq:da}), ii) at each location the branching rate of a given projection is reduced by the density of other projections (Eq.~\ref{eq:comp}), and iii) the branch density of each projection is conserved (Eq.~\ref{eq:norm}).

It is important to emphasize that the formulation of the model is entirely local, insofar as simulation requires no information to be communicated from a given cortical grid cell to any but those immediately adjacent (via diffusion). Hence the simulations demonstrate how a self-organising system, constrained by genetically specified guidance cues, can faithfully reproduce an arrangement of cell aggregates in one neural structure as a topographic map in another.
%
The model also shows that the shape of the barrels may be primarily determined
by the competitive and cooperative interactions between cells relative to the
shape of the cortical field boundary. And that somatotopic map formation does
not require the pre-specification of centre-points by as yet undetermined
additional developmental mechanisms.

>>>>

\begin{abstract}
Brain development relies on an interplay between genetic specification and self-organisation. Striking examples of this relationship can be found in the  somatosensory brainstem, thalamus, and cortex of rats and mice, where the arrangement of the facial whiskers is preserved in the arrangement of cell aggregates to form precise somatotopic maps. We show in simulation how realistic whisker maps can self-organize, by assuming that information is exchanged between adjacent cells only, under the guidance of gene expression gradients. The resulting model provides a simple account of how patterns of gene expression can constrain spontaneous pattern formation to faithfully reproduce functional maps in subsequent brain structures.
\end{abstract}

@@

\dropcap{S}patial patterns in neural connectivity provide clues about the constraints under which brains evolve and develop \citep{purves_iterated_1992}. Perhaps the most distinctive pattern can be found in the barrel cortex of many rodent species \cite{woolsey_structural_1970}. The barrels are identifiable soon after birth in layer 4 of primary somatosensory cortex as dense clusters of thalamocortical axons, which are enclosed by borders a few neurons thick from postnatal day 3 \citep{erzurumlu_development_2012}. In the plane tangential to the cortical surface the barrels constitute a somatotopic map of the whiskers, with cells within adjacent barrels responding most strongly and quickly to deflection of adjacent whiskers \citep{armstrong-james_flow_1992}. Barrel patterning reflects subcortical whisker maps comprising cell aggregates called barrelettes in the brainstem and barreloids in the thalamus.
%\cite{van_der_loos_barreloids_1976}.

@@

The present results suggest that the key requirements for the emergence of realistic barrel patterning are i) at each cortical location thalamocortical projections compete for a limited number of available synaptic connections (Eqs.~\ref{eq:dc}--\ref{eq:da}), ii) at each location the branching rate of a given projection is reduced by the density of other projections (Eq.~\ref{eq:comp}), and iii) the branch density of each projection is conserved over time (Eq.~\ref{eq:norm}).

It is important to emphasize that the formulation of the model is entirely local, insofar as simulation requires no information to be communicated from a given cortical grid cell to any but those immediately adjacent (via diffusion). Hence the simulations demonstrate how a self-organising system, constrained by genetically specified guidance cues and by the shape of the cortical field boundary, can faithfully reproduce an arrangement of cell aggregates in one neural structure as a topographic map in another.

The emergence of barrels in simulation required competition between thalamic projections in terms of synaptic connectivity, and also in terms of cortical space (as represented by $\chi$). This requirement may account for the absence of barrels in rodents with larger brains, such as capybara, for which competition for space is presumably weaker \citep{Woolsey1975}. Hence, irrespective of whether barrels are necessary for adaptive whisker function, the emergence of somatotopically ordered modular structures may be an inevitable consequence of local competition for cortical territory driven by input from an array of discrete sensory organs \citep{purves_iterated_1992}.

Moreover, the present results confirm that such structures can emerge without requiring the pre-specification of centre-points by as yet undetermined additional developmental mechanisms.

%The model also shows that the shape of the barrels may be primarily determined by the competitive and cooperative interactions between cells relative to the shape of the cortical field boundary. And that somatotopic map formation does not require the pre-specification of centre-points by as yet undetermined additional developmental mechanisms.
